---
title: "Reproducible Research - Course Project 1"
author: "Melissa Sacevich"
date: "4/13/2020"
output: 
md_document: 
variant: markdown_github
---

## Introduction

It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the “quantified self” movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

## Data
The data for this assignment can be downloaded from the course web site:

Dataset: [http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip]

The variables included in this dataset are:

* **steps**: Number of steps taking in a 5-minute interval (missing values are coded as NA) 
* **date**: The date on which the measurement was taken in YYYY-MM-DD format
* **interval**: Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.


## Loading and preprocessing the data
```{r}
#Read data from your working directory
setwd("C:/Users/msacevich/Documents/Rstudio")
ActivityData <- read.csv("activity.csv", header=TRUE)

#Missing values
ActivityDataNulls <- ActivityData[is.na(ActivityData$steps),]

#Data without any missing values
ActivityData1 <- ActivityData[!is.na(ActivityData$steps),]
```

## What is mean total number of steps taken per day?

### 1. Calculate the total number of steps taken per day
```{r}
steps_per_day <- aggregate(steps ~ date, ActivityData1, sum)
```

### 2. Make a histogram of the total number of steps taken each day
```{r}
hist(steps_per_day$steps, main = "Total number of steps per day", xlab = "Steps per day", col = "light blue")
```

### 3. Calculate and report the mean and median of the total number of steps taken per day
```{r}
mean(steps_per_day$steps)
median(steps_per_day$steps)
```

## What is the average daily activity pattern?

### 1. Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r}
steps_by_interval <- aggregate(steps ~ interval, ActivityData1, mean)
plot(steps_by_interval$interval, steps_by_interval$steps, type='l', col=1, main="Average number of steps by Interval", xlab="Time Intervals", ylab="Average number of steps")
```

### 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
max <- steps_by_interval[which.max(steps_by_interval$steps),1]
print(max)
```


## Imputing missing values
### 1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r}
Nulls <- sum(!complete.cases(ActivityData))
print(Nulls)
```

### 2,3. Devise a strategy for filling in all of the missing values in the dataset.Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
mean_steps <- with(ActivityData1, tapply(steps, ActivityData1$interval, mean))
ActivityDataNulls$steps <- mean_steps

NewActivityData <- rbind(ActivityData1, ActivityDataNulls)
```

### 4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day.
```{r}
steps_per_day2 <- aggregate(steps ~ date, NewActivityData, sum)
hist(steps_per_day2$steps, main = "Total number of steps per day", xlab = "Steps per day", col = "blueviolet")

mean(steps_per_day2$steps)
median(steps_per_day2$steps)
```

## Are there differences in activity patterns between weekdays and weekends?

### 1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
```{r}
NewActivityData$day <- "weekday"
NewActivityData$day[weekdays(as.Date(NewActivityData$date), abb=T) %in% c("Sat","Sun")] <- "weekend"
table(NewActivityData$day)
```

### 2. Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).
```{r}
library(lattice)
meanstepsbyday <- aggregate(steps ~ interval + day, data=NewActivityData, FUN="mean")
xyplot(steps ~ interval | day, data=meanstepsbyday, type="l", grid=T, layout=c(1,2), ylab="Number of steps", xlab="5-min. intervals from midnight", main="Average  5-min. activity intervals: Weekdays vs. Weekends")
```
